A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-23 tidyindex.qmd
caption for fig 7

```{r fig-idx-tour, fig.align='center', fig.height=5} 
#| fig-cap: "Six frames selected to explore how varying the weights of the politics dimension changes the index values and country rankings in Global Gender Gap Index (GGGI). The top panel shows the GGGI value against the country, ranked by its original index value in Frame 12. The bottom panel displays the weight used to produce the index values in the top panel, with each frame corresponds to a set of weights. Countries selected includes 1) top-ranked countries with GGGI > 0.85, 2) countries ranked between 57 and 62 with GGGI from 0.72 to 0.73, and 3) low-ranked countries with GGGI < 0.6. Compared to Frame 12 where equal weigts are used for the four dimension, a reduced weight in politics (Frame 1 and 6) shows narrower gaps between top and mid- or lower- ranked countries, while an increase in the politics weight (Frame 18, 24, and 29) leads to a systematic decrease of GGGI values across all the countries, except for Bangladesh. Full animation is available at https://vimeo.com/847874016."  
#frames <- c("001", "013", "037", "061", "085", "109")
frames <- c("001", "006", "012", "018", "024", "029")
ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl, ncol = 3)  
```

@@

Errors in measurement, variability and sampling error, may arise at various stages of the pipeline calculation, including from different parameterisation choices, as illustrated from @sec-example1, or from the statistical summarization procedures applied in the pipeline. Although it may not be possible to perfectly measure these errors, it is important that they are recognised and estimated for an index, especially to provide potential confidence intervals. In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate one possibility to compute a confidence interval for the Standardized Precipitation Index (SPI). Bootstrapping is used to account for the sampling uncertainty in the distribution fit step of the index pipeline and to assess its impact on the SPI series.

>>>>

```{r fig-idx-tour, fig.align='center', fig.height=5} 
#| fig-cap: "Exploring the sensitivity of the GGGI, by varying the politics component's contribution, for a subset of countries. Each panel shows a dotplot of the index values, computed for the linear combination represented by the segment plots below. Frame 12 shows the actual GGGI values, and countries are sorted from highest to lowest on this. Frames 1 and 6 show the GGGI if the politics component is reduced. Frames 18, 24, 29 show the GGGI when the politics component is increased. The most notable feature is that Bangladesh's GGGI drops substantially when politics is removed, indicating that this component plays a large role in it's relatively high value. Also, politics plays a substantial role in the GGGI's for the top ranked countries, because each of them drops, to the state of being similar to the middle ranked countries when the politics component's contribution is reduced. The animation can be viewed at https://vimeo.com/847874016."  
#frames <- c("001", "013", "037", "061", "085", "109")
frames <- c("001", "006", "012", "018", "024", "029")
ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl, ncol = 3)  
```

@@

Errors in measurement, variability and sampling error, may arise at various stages of the pipeline calculation, including from different parameterization choices, as illustrated from @sec-example1, or from the statistical summarization procedures applied in the pipeline. Although it may not be possible to perfectly measure these errors, it is important that they are recognised and estimated for an index, so that it is possible to compute confidence intervals. In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate one possibility to compute a confidence interval for the Standardized Precipitation Index (SPI). Bootstrapping is used to account for the sampling uncertainty in the distribution fit step of the index pipeline and to assess its impact on the SPI series.

### 2023-08-23 tidyindex.tex
caption for fig 7

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\caption{\label{fig-idx-tour}Six frames selected to explore how varying
the weights of the politics dimension changes the index values and
country rankings in Global Gender Gap Index (GGGI). The top panel shows
the GGGI value against the country, ranked by its original index value
in Frame 12. The bottom panel displays the weight used to produce the
index values in the top panel, with each frame corresponds to a set of
weights. Countries selected includes 1) top-ranked countries with GGGI
\textgreater{} 0.85, 2) countries ranked between 57 and 62 with GGGI
from 0.72 to 0.73, and 3) low-ranked countries with GGGI \textless{}
0.6. Compared to Frame 12 where equal weigts are used for the four
dimension, a reduced weight in politics (Frame 1 and 6) shows narrower
gaps between top and mid- or lower- ranked countries, while an increase
in the politics weight (Frame 18, 24, and 29) leads to a systematic
decrease of GGGI values across all the countries, except for Bangladesh.
Full animation is available at https://vimeo.com/847874016.}

@@

Errors in measurement, variability and sampling error, may arise at
various stages of the pipeline calculation, including from different
parameterisation choices, as illustrated from
Section~\ref{sec-example1}, or from the statistical summarization
procedures applied in the pipeline. Although it may not be possible to
perfectly measure these errors, it is important that they are recognised
and estimated for an index, especially to provide potential confidence
intervals. In this example, the Texas post office station highlighted in
Figure~\ref{fig-compute-temporal} is used to illustrate one possibility
to compute a confidence interval for the Standardized Precipitation
Index (SPI). Bootstrapping is used to account for the sampling
uncertainty in the distribution fit step of the index pipeline and to
assess its impact on the SPI series.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi

@@

\caption{\label{fig-idx-tour}Exploring the sensitivity of the GGGI, by
varying the politics component's contribution, for a subset of
countries. Each panel shows a dotplot of the index values, computed for
the linear combination represented by the segment plots below. Frame 12
shows the actual GGGI values, and countries are sorted from highest to
lowest on this. Frames 1 and 6 show the GGGI if the politics component
is reduced. Frames 18, 24, 29 show the GGGI when the politics component
is increased. The most notable feature is that Bangladesh's GGGI drops
substantially when politics is removed, indicating that this component
plays a large role in it's relatively high value. Also, politics plays a
substantial role in the GGGI's for the top ranked countries, because
each of them drops, to the state of being similar to the middle ranked
countries when the politics component's contribution is reduced. The
animation can be viewed at https://vimeo.com/847874016.}

@@

Errors in measurement, variability and sampling error, may arise at
various stages of the pipeline calculation, including from different
parameterization choices, as illustrated from
Section~\ref{sec-example1}, or from the statistical summarization
procedures applied in the pipeline. Although it may not be possible to
perfectly measure these errors, it is important that they are recognised
and estimated for an index, so that it is possible to compute confidence
intervals. In this example, the Texas post office station highlighted in
Figure~\ref{fig-compute-temporal} is used to illustrate one possibility
to compute a confidence interval for the Standardized Precipitation
Index (SPI). Bootstrapping is used to account for the sampling
uncertainty in the distribution fit step of the index pipeline and to
assess its impact on the SPI series.
